Extended Data

### 2021-11-18 doc/manuscript/Dow_manuscript_Nature.Rmd
add GH ORCID
GLH feedback

Grant L. Harley^5^

@@

5. Department of Geography & Geological Sciences, University of Idaho, ID 83844, USA

@@

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014].
Current models assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
However, recent experimental and observational findings that annual productivity can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021], and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018], suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.

While the responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], very little is known about how the longest-lived component of fixed C in trees, the woody growth, is responding to warming spring temperatures. 
In fact, we are aware of only one study that has documented stem-growth phenology of temperate deciduous species over multiple years[@dorangeville_peak_2021]. <!-- I'm not aware of any others, but we do want to make sure we're not missing anything. klesse_integrating_2016 and martin_hydrometeorology_2017 both do this, but for conifers.-->
The climate sensitivity of woody growth phenology in temperate deciduous trees and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a controlled sapling experiment).

Tree-ring records, which can be used to examine correlations of annual growth to temperature but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019],
<!-- dorangeville_drought_2018 looks at PET.-->
with some hints that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019].
<!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
Thus, tree-ring evidence does not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake[@keenan_net_2014].
Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.

@@

Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year when 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), the maximum daily growth rates ($g_{max}$) and the DOY on which it occurred ($DOY_{g_{max}}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@dorangeville_peak_2021]. These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).

@@

## Tree ring analysis {-#Tree}

To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies of `r n_chronologies_SCBI` species at SCBI[@helcoski_growing_2019] and `r n_chronologies_HF` species at Harvard Forest (`r EDT_species`), along with an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig. `r tree_rings`; `r EDF_map`; `r EDT_tree_ring_summary`)[@dorangeville_drought_2018].
In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in April $T_{max}$.
Across all chronologies, RWI was significantly (at p $\le$ 0.05) positively correlated with April $T_{max}$ for only 2% of chronologies: 1 of 142 ring-porous and 4 of 66 diffuse-porous species-site combinations (`r EDT_tree_ring_summary`).
In contrast, RWI was frequently negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 52% (May: 45/141, Jun: 107/141, Jul: 91/141, Aug: 53/141) and 46% (May: 10/66, Jun: 52/66, Jul: 36/66, Aug: 23/66) of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
$T_{min}$ generally exibited weaker relationships to annual growth than $T_{max}$, with few significant correlations between spring $T_{min}$ and RWI (`r EDF_tree_rings_TMIN`).

@@

One possibility is that extra photosynthate in years with warm springs may be allocated to woody growth without affecting diameter growth in the current year. 
It is theoretically possible that extra C is allocated to cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a thicker layer of higher-density latewood, resulting in formation of more C-dense wood in years with warm springs.
However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006; @roibu_climatic_2020; @kern_multiple_2013], which is more strongly controlled by summer drought stress[@tardif_influence_2006; @roibu_climatic_2020], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely. 
Extra C could also be saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013], potentially including an earlier start to growth[@fu_variation_2014]. 
Extension of our tree-ring analysis revealed little correlation between April $T_{max}$ and growth the following year (sig. pos. correlations for 5/142 RP and 3/66 DP species-site combinations, Fig. `r EDF_tree_rings_TMAX_extended`), although predominantly positive (non-significant) correlations in RP species suggests that this dynamic may weakly influence their annual growth.
Thus, warm springs are unlikely to provide substantial, sustained C sinks under warming spring temperatures.

@@

Dendrometer band measurements were collected at SCBI[@bourg_initial_2013] and Harvard Forest[@finzi_carbon_2020;@dorangeville_peak_2021], both part of the Forest Global Earth Observatory (ForestGEO)[@anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021]. 
SCBI (38.8935° N, 78.1454° W; elevation 273–338 m.a.s.l.) is located in the Blue Ridge Mountains at the northern end of Shenandoah National Park, 5 km South of Front Royal, Virginia. 
The forest is secondary and mixed age, having established in the mid-19th century after conversion from agricultural fields[@bourg_initial_2013]. 
Dominant canopy species within the 25.6 ha ForestGEO plot include tulip poplar (*Liriodendron tulipifera* L.), oaks (*Quercus spp.*), and hickories (*Carya spp.*)[@helcoski_growing_2019]. 
The climate is humid temperate, with 1950-2019 mean annual precipitation of 1018 mm and temperatures averaging 1° C in January and 24° C in July[@anderson-teixeira_joint_2021].

@@

For each species-site combination, we converted tree ring records into a dimensionless ring width index (RWI) to emphasize interannual variability associated with climate [@cook_methods_1990].
A 2/3rds _n_ spline was applied to each core using ARSTAN to produce standardized ring-width series; _n_ is the number of years in each series[@cook_time_1985; @cook_methods_1990].
An adaptive power transformation, a process that also stabilises the variance over time[@cook_calculating_1997], was used to minimize the influence of outliers in all series. 
Low series replication, often in the earliest portions of a chronology collection, can also inflate the variance of tree-ring records[@jones_estimating_1997].
The 1/3rds spline method was chosen when replication in the inner portion of each chronology (ca. inner 30–50 yr of each record depending on full chronology length) was less than three trees. 
When replication was greater than that, we used the average correlation between raw ring-width series (rbar) method.
The robust biweight mean chronology (RWI) for each species-site combination was calculated from the ring-width indices following variance stabilisation[@cook_time_1985].
We defined chronology start year (Extended Data Table 1) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.

<!--QUILT PLOT METHODS-->
For the analysis of correlation between RWI and climate variables, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
Correlations between monthly climate and $RWI$ were assessed using ‘dplR’[@bunn_dendrochronology_2008] and ‘bootRes’[@zang_dendroclimatic_2013] in R v 4.0 (R Core Team, 2020), which correlated functions and bootstrapped confidence intervals for these relationships[@biondi_dendroclim2002_2004].
We analyzed these correlations for January through September of the current year (presented in Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
To test for potential lag effects of spring temperatures on the following year's growth, we also ran a version of the analysis extending back to include climate of every month of the previous year (`r EDF_tree_rings_TMAX_extended`).
Correlations and significance levels for months April-August are given in `r chron_table`.

>>>>

Grant L. Harley^5^ ( Orcid ID : 0000-0003-1557-8465)

@@

5. Department of Earth and Spatial Sciences, University of Idaho, ID 83844, USA

@@

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014].
Current models assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
However, recent experimental and observational findings show that annual productivity can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021], and that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018]. These studies suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.

While some studies addressing the responses of leaf phenology and seasonal _NEE_ to warming spring temperatures exist[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody growth, is responding to warming spring temperatures. 
In fact, we are aware of only one study that has documented stem-growth phenology of temperate deciduous species over multiple years[@dorangeville_peak_2021]. <!-- I'm not aware of any others, but we do want to make sure we're not missing anything. klesse_integrating_2016 and martin_hydrometeorology_2017 both do this, but for conifers.-->
The climate sensitivity of woody growth phenology in temperate deciduous trees and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a controlled sapling experiment).

Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019],
<!-- dorangeville_drought_2018 looks at PET.-->
with some hints that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019].
<!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
Thus, tree-ring evidence does not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake[@keenan_net_2014].
Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.

@@

Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), the maximum daily growth rates ($g_{max}$) and the DOY on which it occurred ($DOY_{g_{max}}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@dorangeville_peak_2021]. These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).

@@

## Tree-ring analysis {-#Tree}

To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies of `r n_chronologies_SCBI` species at SCBI[@helcoski_growing_2019] and `r n_chronologies_HF` species at Harvard Forest (`r EDT_species`), along with an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig. `r tree_rings`; `r EDF_map`; `r EDT_tree_ring_summary`)[@dorangeville_drought_2018].
In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in April $T_{max}$.
Across all chronologies, the standardized ring-width index (RWI) was significantly (at p $\le$ 0.05) positively correlated with April $T_{max}$ for only 2% of chronologies: 1 of 142 ring-porous and 4 of 66 diffuse-porous species-site combinations (`r EDT_tree_ring_summary`).
In contrast, RWI was frequently negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 52% (May: 45/141, Jun: 107/141, Jul: 91/141, Aug: 53/141) and 46% (May: 10/66, Jun: 52/66, Jul: 36/66, Aug: 23/66) of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
$T_{min}$ generally exibited weaker relationships to annual growth than $T_{max}$, with few significant correlations between spring $T_{min}$ and RWI (`r EDF_tree_rings_TMIN`).

@@

One possibility is that extra photosynthate in years with warm springs may be allocated to woody growth without affecting diameter growth in the current year. 
It is theoretically possible that extra C is allocated to cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a thicker layer of higher-density latewood, resulting in formation of more C-dense wood in years with warm springs.
However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006; @roibu_climatic_2020; @kern_multiple_2013], which is more strongly controlled by summer drought stress[@tardif_influence_2006; @roibu_climatic_2020], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely. 
Extra C could also be saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013], potentially including an earlier start to growth[@fu_variation_2014]. 
Extension of our tree-ring analysis revealed weak correlation between April $T_{max}$ and growth the following year (sig. pos. correlations for 5/142 RP and 3/66 DP species-site combinations, Fig. `r EDF_tree_rings_TMAX_extended`), although predominantly positive (non-significant) correlations in RP species suggests that this dynamic may weakly influence their annual growth.
Thus, warm springs are unlikely to provide substantial, sustained C sinks under warming spring temperatures.

@@

Dendrometer band measurements were collected at SCBI[@bourg_initial_2013] and Harvard Forest[@finzi_carbon_2020;@dorangeville_peak_2021], both part of the Forest Global Earth Observatory (ForestGEO)[@anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021]. 
SCBI (38.8935° N, 78.1454° W; elevation 273–338 m.a.s.l.) is located in the Blue Ridge Mountains at the northern end of Shenandoah National Park, 5 km south of Front Royal, Virginia. 
The forest is secondary and mixed age, having established in the mid-19th century after conversion from agricultural fields[@bourg_initial_2013]. 
Dominant canopy species within the 25.6 ha ForestGEO plot include tulip poplar (*Liriodendron tulipifera* L.), oaks (*Quercus spp.*), and hickories (*Carya spp.*)[@helcoski_growing_2019]. 
The climate is humid temperate, with 1950-2019 mean annual precipitation of 1018 mm and temperatures averaging 1° C in January and 24° C in July[@anderson-teixeira_joint_2021].

@@

For each species-site combination, we converted tree-ring records into the dimensionless RWI to emphasize interannual variability associated with climate [@cook_methods_1990].
A 2/3rds _n_ spline was applied to each core using ARSTAN to produce standardized ring-width series; _n_ is the number of years in each series[@cook_time_1985; @cook_methods_1990].
An adaptive power transformation, a process that also stabilises the variance over time[@cook_calculating_1997], was used to minimize the influence of outliers in all series. 
Low series replication, often in the earliest portions of a chronology collection, can also inflate the variance of tree-ring records[@jones_estimating_1997].
The 1/3rds spline method was chosen when replication in the inner portion of each chronology (ca. inner 30–50 yr of each record depending on full chronology length) was less than three trees. 
When replication was greater than n = 3 trees, we used the average correlation between raw ring-width series (rbar) method.
The robust biweight mean chronology (RWI) for each species-site combination was calculated from the ring-width indices following variance stabilisation[@cook_time_1985].
We defined chronology start year (Extended Data Table 1) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.

<!--QUILT PLOT METHODS-->
For the analysis of correlation between RWI and climate variables, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
Correlations between monthly climate and $RWI$ were assessed using ‘dplR’[@bunn_dendrochronology_2008] and ‘bootRes’[@zang_dendroclimatic_2013] in R v 4.0 (R Core Team, 2020), which correlated functions and bootstrapped confidence intervals for these relationships[@biondi_dendroclim2002_2004].
We analyzed these correlations for January through September of the current year (presented in Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
To test for potential lag effects of spring temperatures on growth the following year, we also ran a version of the analysis extending back to include climate of every month of the previous year (`r EDF_tree_rings_TMAX_extended`).
Correlations and significance levels for months April-August are given in `r chron_table`.
